A Speculative Approach to Parallelization in Particle Swarm Optimization

### 2010-10-08 honors_thesis/si_paper/paper.tex
Done with the SI paper, and submitted it

\documentclass[journal,letterpaper]{IEEEtran}

\usepackage{cite}
\usepackage{graphicx}
\usepackage{amsmath,amssymb}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{psfrag}

\IEEEoverridecommandlockouts

@@

\title{\ \\ \LARGE\bf A Speculative Approach to Parallelization in Particle
Swarm Optimization%
\thanks{Matthew Gardner, Andrew McNabb, and Kevin Seppi are with the Department
of Computer Science, Brigham Young University, 3361 TMCB, Provo, UT 84602
(phone: 801-422-8717; email: \{mjg82,a,k\}@cs.byu.edu).}%
}

\date{}

@@

@@

@@

\begin{keywords}
Parallel algorithms, optimization methods, particle swarm optimization.
\end{keywords}

@@

Within the last few years, researchers have begun to recognize the need to
develop parallel implementations of PSO, publishing many papers on the subject.
The methods they have used include various synchronous algorithms
\cite{chu-2006-intelligent-parallel-pso, jin-2005-pso-antenna-designs,
parsopoulos-2004-parallel-vector-evaluated-pso,
schutte-2004-parallel-global-optimization-with-pso} and asynchronous algorithms
\cite{mostaghim-2006-multi-objective-pso-on-grids,
venter-2005-parallel-pso-asynchronous-evaluations}.  Parallelizing the
evaluation of the objective function can also be done in some cases, though
that is not an adaption of the PSO algorithm itself and thus is not the focus
of this paper.

These previous parallel techniques distribute the computation needed by the
particles in the swarm over the available processors.  If more processors are
available, these techniques increase the number of particles in the swarm,
either by adding individual particles or by adding entire new sub-swarms.  In
almost all cases, adding additional particles produces better results in the
same amount of time~\cite{mcnabb-2009-large-particle-swarms}.  In
\fig{iters-sphere} we see an example of this on the well-known benchmark
function Sphere (20 dimensions, reporting the average of twenty runs).  In
terms of the number of iterations performed (which is equivalent to wall-clock
time if all particles are evaluated in parallel), every time the swarm size
increases, the performance improves.

@@

However, it can be seen from the graph that doubling the swarm size produces
only incremental benefit; performance improves, but not by very much.  The
reason for this is that there comes a point of diminishing returns with respect
to adding particles.  In \fig{evals-sphere} we show the value obtained after
50,000 function evaluations (not iterations) as a function of swarm size, again
for the function Sphere.  Increasing the swarm size from 5 to 10 has a
significant effect on the value obtained.  However, increasing the swarm size
from 16 to 30 makes the algorithm less efficient; that is, it reduces the
progress the algorithm makes per evaluation.  Other functions show similar
trends, though often the optimal swarm size is slightly larger.  For this
reason, previous work has recommended the use of a swarm size of 50 for
PSO~\cite{bratton-2007-defining-a-standard-for-pso}.  Thus, in at least some
cases, adding particles indefinitely will not yield an efficient
implementation.

@@

To solve this problem, we apply the concept of speculative
decomposition~\cite{grama-2003-intro-to-parallel-computing} to particle swarm
optimization, using extra processors to perform two iterations of PSO at the
same time.  Speculative decomposition is analogous to speculative execution
(also known as branch prediction), a technique commonly used in processors.
Modern processors, when faced with a branch on which they must wait (e.g., a
memory cache miss), guess which way the branch will go and start executing,
ensuring that any changes can be undone.  If the processor guesses right,
execution is much farther ahead than if it had idly waited on the memory
reference.  If it guesses wrong, execution restarts where it would have been
anyway.

@@

Furthermore, we show that if we relax the requirements of the algorithm, no
longer demanding that it strictly reproduce the exact behavior of standard PSO,
we can introduce new speculative techniques that often out-perform PSO.  These
relaxations make better use of the information obtained from the extra
exploration made by the speculative function evaluations.  We also explore the
idea that, like branch prediction in processors, we need not speculatively
evaluate \emph{all} possible future positions, we can accelerate the algorithm
even if we are just \emph{likely} to have guessed right.  By pruning the
speculation to just paths that are statistically likely to reproduce the paths
that are equivalent to PSO we can increase the swarm size without increasing
the number of speculative evaluations.  We also consider several recovery
strategies for cases where the pruned set of speculative evaluations does not
contain the evaluation that standard PSO would have done.  A further
improvement we explore is speculating several iterations ahead instead of just
one, which is made possible by pruning the number of speculative evaluations.

@@

Particle swarm optimization was proposed in 1995 by James Kennedy and Russell
Eberhart~\cite{kennedy-1995-particle-swarm-optimization}.  The algorithm is
used to intelligently search a multi-dimensional space by mimicking the
swarming and flocking behavior of birds and other animals. It is a social
algorithm that depends on interaction between particles to quickly and
consistently approximate the optimal solution to a given objective function.

@@

At each iteration of the algorithm, the position $\ppos_t$ and velocity
$\pvel_t$ of each particle are updated as follows:
\begin{align}
\nonumber
	\pvel_{t+1} &=
		\constriction \bigl[ \pvel_t
			+ \pcoeff\pURand_{t}\otimes(\pbest_{t} - \ppos_{t}) \\
\label{eq:velupdate}
			& \quad \quad \quad \, + \ncoeff\nURand_{t}\otimes(\nbest_{t} - \ppos_{t})
		\bigr] \\
\label{eq:posupdate}
	\ppos_{t+1} &= \ppos_{t} + \pvel_{t+1}
\end{align}
where \( \pURand_{t} \) and \( \nURand_{t} \) are vectors of independent random
numbers drawn from a standard uniform distribution, the \( \otimes \) operator
is an element-wise vector multiplication, $\pbest$ (called personal best) is
the best position the current particle has seen, and $\nbest$ (called
neighborhood best) is the best position the neighbors of the current particle
have seen~\cite{bratton-2007-defining-a-standard-for-pso}.  The parameters \(
\ncoeff \), \( \pcoeff \), and \( \constriction \) are given prescribed values
required to ensure convergence (2.05, 2.05, and .73,
respectively)~\cite{clerc-2002-constricted-pso}.

Changing the way neighbors are defined, usually called the ``topology,'' has a
significant effect on the performance of the algorithm.  In the Ring topology,
each particle has one neighbor to either side of it; in the Complete
topology\footnote{The Complete topology has often been unfortunately named Star
in the literature, which in graph theory refers to a completely different
topology.  Other names have also been used, including ``global topology'' and
gbest.  We use the graph theory term ``Complete'' in this paper.}, every
particle is a neighbor to every other
particle~\cite{bratton-2007-defining-a-standard-for-pso}.  In all topologies a
particle is also a neighbor to itself in that its own position and value are
considered when updating the particle's neighborhood best, $\nbest$.  Thus with
$p$ particles, using the Ring topology each particle with index $i$ has three
neighbors: $i-1$, $i$ (itself), and $i+1$.  With the Complete topology, each
particle has $p$ neighbors.

In this paper we use these topologies as well as a parallel adaptation of the
Complete topology, called Random, that has been shown to approximate the
behavior of Complete with far less
communication~\cite{mcnabb-2009-large-particle-swarms}.  In the Random
topology, each particle randomly picks two other particles to share information
with at each iteration, along with itself.  Thus in both the Ring and the
Random topologies, all particles have three neighbors.

@@

Because the number of speculative evaluations depends on the number of
neighbors a particle has, the choice of topology is an important one.  The use
of the Complete topology, where every particle is a neighbor to every other
particle, would require $O(p^2)$ speculative evaluations per iteration.
Clearly it is much more desirable to have a sparse topology, where $O(np)$ is
much smaller than $O(p^2)$.  However, some functions are better optimized with
the Complete topology and the quick spread of information it entails than with
sparse topologies.  Accordingly, we use the Random topology described
in~\cite{mcnabb-2009-large-particle-swarms}, which has been shown to
effectively simulate the Complete topology.  In~\sec{results} we report the
results for performing speculative evaluation using both the Ring topology and
the Random topology on a number of common benchmark functions.

@@

In a centralized, or Master-Slave, parallel PSO algorithm, one machine, the
master, keeps track of all necessary information, and all other machines are
merely used to evaluate the objective function at various positions as directed
by the master~\cite{belal-2004-parallel-models-for-pso}.  To perform
speculative evaluation in such an architecture, the master generates the
positions to evaluate speculatively as in \eqref{eq:defpcasegen}.  After having
the slaves evaluate the objective function at all necessary positions, the
master then decides which position to accept for each particle, as in
\eqref{eq:val2updatelong}.  The outline of the procedure is given in
\alg{centralized}.

@@

In a distributed parallel PSO algorithm, individual processors not only perform
evaluations of particles, but also their movement.  The information for each
particle is not held by a central machine that directs the algorithm; instead,
each processor has the information for the particle or particles that it is in
charge of and must perform the steps of the algorithm for those
particles~\cite{mcnabb-2007-parallel-pso-using-mapreduce}.  Messages such as
values and positions for the neighborhood best are sent between processors.
There may still be some machine that collects information from all of the
particles and outputs the result of the algorithm, though that machine's
importance is much less than in centralized algorithms.

@@

While the idea of speculative evaluation in particle swarm optimization is
intriguing, the results in \sec{results} show that in many cases it requires
too many extra evaluations to produce competitive results.  On most functions,
the larger swarm size that can be used in regular PSO leads to better
performance than is found by trying to do speculative evaluation.  However, if
we keep the idea of speculative evaluation while relaxing the requirement of
exactly reproducing the behavior of the original PSO algorithm, we see some
impressive results.

@@

@@

At this point it is also interesting to note a parallel between our methods and
parallel evolution
strategies~\cite{rudolph-1991-distributed-evolution-strategies}.  In evolution
strategies, a parent individual (representing a potential solution to some
objective function) produces a number of offspring by a mutation operator.  One
of the individuals is selected by a selection operator, and that individual
becomes the parent for the next
generation~\cite{beyer-2002-evolution-strategies}.  Our methods are similar,
where our mutation operator is simply the PSO motion equations and the
selection operator is either the indicator function introduced in \sec{sepso},
in the case of our original speculative algorithm, or the standard selector
operator based on fitness, in the case of this Pick Best technique.

@@

When not all of the branches are evaluated, there is some probability of not
evaluating the branch that was actually taken by the original particle.  There
are a few possibilities for recovery in this case.  One is to leave the
particle as it is, not accepting any of the speculative evaluations, because
none of them were correct.  This leads to particles being at different
iterations, as some particles guess correctly while others do not.  Thus we
lose exact compatibility with the original PSO, though this particular
relaxation is nothing new; asynchronous adaptations of PSO do the same
thing~\cite{koh-2006-parallel-asynchronous-pso}.  As an aside, it is equivalent
in this case to simply increment the iteration number of particles which fail
to correctly predict their branch.  This keeps the iteration number constant
across all particles, simplifying the work needed to be done in determining
neighbors when dynamic topologies are involved.  We call this technique Social
Promotion.

@@

We ran experiments to compare our speculative PSO algorithm to the standard PSO
algorithm.  At each iteration of the algorithms, we use one processor to
perform one function evaluation for one particle, be it speculative or
non-speculative.  The speculative algorithm actually performs two iterations of
PSO at each ``iteration,'' so we instead call each ``iteration'' a ``round of
evaluations.''

Because our interest is in exploring the most efficient use of processors, we
must hold fixed the number of processors used at each round of evaluations in
all of our experiments, giving all algorithms the same number of function
evaluations.  Thus while we label our graphs with ``Rounds of Evaluations,''
this is equivalent to the number of function evaluations used by each algorithm
(simply multiply by the number of processors to transform rounds of evaluations
to function evaluations).  We use ``Rounds of Evaluations'' to reflect
wall-clock time, as our primary interest is using extra processors in parallel
to improve performance on functions with long evaluation times.  A smaller
swarm running serial PSO would likely produce better results than we present in
the same number of function evaluations, but because the function evaluations
were not done in parallel it would take more wall-clock time.  When comparing
the performance of parallelized algorithms on long-running functions, we must
compare by wall-clock time with the same resources, not solely by the number of
function evaluations.

Using the same number of processors for each algorithm requires that our
speculative algorithms have a smaller swarm size than standard PSO.  For each
of the topologies we used, a particle has three neighbors including itself.  As
shown in Table~\ref{tab:evals}, this results in $7$ speculative evaluations per
particle.  With one evaluation needed for the original, non-speculative
particle, we have $8p$ evaluations for every two iterations, where $p$ is the
number of particles in the speculative swarm.  The extra evaluations required
in our speculative approach would instead be used to evaluate particles in
regular PSO, so we compare swarms of size $p$ in speculative evaluation with
swarms of size $8p$ in regular PSO.

Where the Complete topology would normally be preferred, we use a Random
topology in our speculative algorithm, as Complete leads to an explosion in the
number of speculative evaluations.  If speculative evaluation were not being
performed, it is possible that the Complete topology would be used.  However,
the Complete topology also requires a very large amount of interprocessor
communication in distributed PSO, so it is still quite possible that Random
would be used even with standard PSO.  But, to be fair in our comparisons, we
compare to the standard PSO algorithm using both the Random topology and the
Complete topology (labeled PSO Random and PSO Complete in our results).

We experimented with five common benchmark problems defined
in~\cite{herrera-2010-test-suite}: Rastrigin, Sphere, Schwefel 2.21, Griewank,
and Bohachevsky.

@@

The $c$ in the function definitions is a shifted center, in order to avoid
origin-seeking bias in the PSO algorithm and its
variants~\cite{monson-2005-origin-seeking-bias}.  We move the center of the
shifted functions (all except Bohachevsky) to be halfway between the center and
the boundary of the initialization region.  For example, if the initialization
region is $[-50,50]^D$, the center is $(25)^D$.  For all of these functions we
used 20 dimensions.

The algorithms we compare are the standard PSO algorithm, our original
speculative approach (here labeled SEPSO simply for ease of reference), and the
four relaxations of SEPSO discussed in \sec{relax}.  In presenting our results,
we call the approach developed in \sec{pickbest} Pick Best.  The algorithms
described in \sec{pruning} through \sec{wrong} are called Pick Best Pruned and
Social Promotion Pruned, and the algorithm in \sec{manyiters} we call Many
Iterations.

@@

Because pruning only requires two speculative evaluations per particle (along
with evaluating the original particle), we can use swarms of size
$\frac{1}{3}p$ to compare to a swarm of size $p$ with standard PSO, instead of
$\frac{1}{8}p$ with other techniques.

@@

First we look at Rastrigin.  Rastrigin is a multi-modal function that is best
optimized with a large, Complete swarm.  It has been shown that with Rastrigin,
the more particles there are in the swarm, the lower function value it finds,
up to at least 4000 particles~\cite{mcnabb-2009-large-particle-swarms}.
Smaller swarms get caught in local optima.  Because our speculative algorithms
require significantly smaller swarm sizes, we would expect to not perform very
well on functions such as Rastrigin.  Our experiments show our intuition to be
correct.  In this experiment we used 240 processors, so SEPSO and Pick Best
each had 30 particles, and the standard PSO algorithms had 240 particles.  As
expected, SEPSO and Pick Best converge quickly to worse local optima than
standard PSO does.  \fig{rastrigin} shows the results graphically.

@@

Next we go to Sphere, the simplest of common benchmark functions.  The function
has no local optima (only a single global optimum) and is best optimized in
terms of function evaluations with a small swarm using a Complete topology.  We
expect our methods to be perfect for such functions, and our results show this
intuition again to be correct.  First we compare the basic SEPSO algorithm to
standard PSO.  In \fig{basic-sphere} we can see that when both SEPSO and PSO
are using a Random topology, SEPSO clearly outperforms PSO.  However, the
comparison between SEPSO and PSO Complete is not as encouraging---PSO Complete
performs slightly better.

@@

When we introduce Pick Best, we see more impressive results.  As can be seen in
\fig{sphere-pickbest}, where SEPSO did not manage to perform as well as PSO
Complete, Pick Best handily outperforms it with the same number of processors.

@@

Still more impressive with Sphere are the results of pruning and speculating
several iterations ahead.  Pick Best Pruned outperforms Pick Best, and Many
Iterations by far outperforms every other algorithm on this function.  Many
Iterations on average finds values that are more than 40 orders of magnitude
better than standard PSO using the same number of processors.  This is shown in
\fig{sphere-manyiters}.  Social Promotion Pruned performed comparably with
SEPSO, and we do not clutter the graph with a line for it.

@@

Schwefel 2.21 is a function similar to Sphere, but Schwefel 2.21 is benefited 
more by larger swarms than Sphere is.  Thus our speculative algorithms often
fail to perform as well as standard PSO because of their smaller swarm size.
In comparing SEPSO to standard PSO in \fig{schwefel1}, we can see that SEPSO
fails to perform any better than PSO with a Random topology, and it is by far
outperformed by PSO with a Complete topology.

@@

As can be seen in \fig{schwefel2}, almost all of the speculative algorithms
perform worse than PSO Complete, with Many Iterations performing comparably to
PSO.  However, one of the speculative algorithms did perform better than PSO:
Pick Best Pruned.  It seems that the larger swarm size available because of
pruning was the cause of the improved performance.

@@

Because the speculative algorithm got stuck with such a small swarm size, we
ran a set of experiments on Griewank with 800 processors, giving swarms of size
100 and 800 instead of 30 and 240.  \fig{basic-griewank2} shows the results.
\fig{basic-griewank2} is very similar to \fig{basic-griewank1}, except that
speculative evaluation finds the global optimum essentially 100\% of the time.
Thus with a few more processors, our algorithm finds the optimum on average
twice as fast as the original PSO without a significant loss of accuracy.

@@

In order to show the premature convergence properties of the various algorithms
we have presented, we will show results first with swarms using 240 processors,
where these are more visible.  Then we will present an overall result for 800
processors, where using our techniques is generally more successful.

>>>>

\documentclass[smallcondensed]{svjour3}

\usepackage{graphicx}
\usepackage{amsmath,amssymb}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{psfrag}
\usepackage{natbib}

@@

\title{A Speculative Approach to Parallelization in Particle Swarm
Optimization}

\date{Received: date / Accepted: date}

@@

\institute{M. Gardner \at
Brigham Young University.
3342 TMCB, Provo, UT 84604
Tel.: 801-422-8717
Fax: 801-422-0169
\email{mjg82@byu.edu}
\and
A. McNabb \at
Brigham Young University.
3342 TMCB, Provo, UT 84604
\email{a@cs.byu.edu}
\and
K. Seppi \at
Brigham Young University.
3330 TMCB, Provo, UT 84604
\email{k@cs.byu.edu}
}

@@

\keywords{Parallel algorithms \and Optimization methods \and Particle swarm
optimization \and Speculative Decomposition}

@@

@@

Within the last few years, researchers have begun to recognize the need to
develop parallel implementations of PSO, publishing many papers on the subject.
The methods they have used include various synchronous algorithms
\citep{chu-2006-intelligent-parallel-pso,jin-2005-pso-antenna-designs,%
parsopoulos-2004-parallel-vector-evaluated-pso,%
schutte-2004-parallel-global-optimization-with-pso} and asynchronous algorithms
\citep{mostaghim-2006-multi-objective-pso-on-grids,%
venter-2005-parallel-pso-asynchronous-evaluations}.  Parallelizing the
evaluation of the objective function can also be done in some cases, though
that is not an adaption of the PSO algorithm itself and thus is not the focus
of this paper.

These previous parallel techniques distribute the computation needed by the
particles in the swarm over the available processors.  If more processors are
available, these techniques increase the number of particles in the swarm,
either by adding individual particles or by adding entire new sub-swarms.  In
almost all cases, adding additional particles produces better results in the
same amount of time~\citep{mcnabb-2009-large-particle-swarms}.  In
\fig{iters-sphere} we see an example of this on the well-known benchmark
function Sphere (20 dimensions, reporting the average of twenty runs).  In
terms of the number of iterations performed (which is equivalent to wall-clock
time if all particles are evaluated in parallel), every time the swarm size
increases, the performance improves.

@@

However, it can be seen from the graph that doubling the swarm size produces
only incremental benefit; performance improves, but not by very much.  The
reason for this is that there comes a point of diminishing returns with respect
to adding particles.  In \fig{evals-sphere} we show the value obtained after
50,000 function evaluations (not iterations) as a function of swarm size, again
for the function Sphere.  Increasing the swarm size from 5 to 10 has a
significant effect on the value obtained.  However, increasing the swarm size
from 16 to 30 makes the algorithm less efficient; that is, it reduces the
progress the algorithm makes per evaluation.  Other functions show similar
trends, though often the optimal swarm size is slightly larger.  For this
reason, previous work has recommended the use of a swarm size of 50 for
PSO~\citep{bratton-2007-defining-a-standard-for-pso}.  Thus, in at least some
cases, adding particles indefinitely will not yield an efficient
implementation.

@@

In order to solve this problem, we apply the concept of speculative
decomposition~\citep{grama-2003-intro-to-parallel-computing} to particle swarm
optimization, using extra processors to perform two iterations of PSO at the
same time.  Speculative decomposition is analogous to speculative execution
(also known as branch prediction), a technique commonly used in processors.
Modern processors, when faced with a branch on which they must wait (e.g., a
memory cache miss), guess which way the branch will go and start executing,
ensuring that any changes can be undone.  If the processor guesses right,
execution is much farther ahead than if it had idly waited on the memory
reference.  If it guesses wrong, execution restarts where it would have been
anyway.

@@

Furthermore, we show that if we relax the requirements of the algorithm, no
longer demanding that it strictly reproduce the exact behavior of standard PSO,
we can introduce new speculative techniques that often out-perform na\"ive
parallelizations of PSO.  These relaxations make better use of the information
obtained from the extra exploration made by the speculative function
evaluations.  We also explore the idea that, like branch prediction in
processors, we need not speculatively evaluate \emph{all} possible future
positions, we can accelerate the algorithm even if we are just \emph{likely} to
have guessed right.  By pruning the speculation to just paths that are
statistically likely to reproduce the paths that are equivalent to PSO we can
increase the swarm size without increasing the number of speculative
evaluations.  We also consider several recovery strategies for cases where the
pruned set of speculative evaluations does not contain the evaluation that
standard PSO would have done.  A further improvement we explore is speculating
several iterations ahead instead of just one, which is made possible by pruning
the number of speculative evaluations.

@@

Particle swarm optimization was proposed in 1995 by James Kennedy and Russell
Eberhart~\citep{kennedy-1995-particle-swarm-optimization}.  The algorithm is
used to intelligently search a multi-dimensional space by mimicking the
swarming and flocking behavior of birds and other animals. It is a social
algorithm that depends on interaction between particles to quickly and
consistently approximate the optimal solution to a given objective function.

@@

At each iteration of the algorithm, the position $\ppos_t$ and velocity
$\pvel_t$ of each particle are updated as follows:
\begin{align}
\nonumber
	\pvel_{t+1} &=
		\constriction \bigl[ \pvel_t
			+ \pcoeff\pURand_{t}\otimes(\pbest_{t} - \ppos_{t}) \\
\label{eq:velupdate}
			& \quad \quad \quad \, + \ncoeff\nURand_{t}\otimes(\nbest_{t} - \ppos_{t})
		\bigr] \\
\label{eq:posupdate}
	\ppos_{t+1} &= \ppos_{t} + \pvel_{t+1}
\end{align}
where \( \pURand_{t} \) and \( \nURand_{t} \) are vectors of independent random
numbers drawn from a standard uniform distribution, the \( \otimes \) operator
is an element-wise vector multiplication, $\pbest$ (called personal best) is
the best position the current particle has seen, and $\nbest$ (called
neighborhood best) is the best position the neighbors of the current particle
have seen~\citep{bratton-2007-defining-a-standard-for-pso}.  The parameters \(
\ncoeff \), \( \pcoeff \), and \( \constriction \) are given prescribed values
required to ensure convergence (2.05, 2.05, and .73,
respectively)~\citep{clerc-2002-constricted-pso}.

Changing the way neighbors are defined, usually called the ``topology,'' has a
significant effect on the performance of the algorithm.  In the Ring topology,
each particle has one neighbor to either side of it; in the Complete
topology\footnote{The Complete topology has often been unfortunately named Star
in the literature, which in graph theory refers to a completely different
topology.  Other names have also been used, including ``global topology'' and
gbest.  We use the graph theory term ``Complete'' in this paper.}, every
particle is a neighbor to every other
particle~\citep{bratton-2007-defining-a-standard-for-pso}.  In all topologies a
particle is also a neighbor to itself in that its own position and value are
considered when updating the particle's neighborhood best, $\nbest$.  Thus with
$p$ particles, using the Ring topology each particle with index $i$ has three
neighbors: $i-1$, $i$ (itself), and $i+1$.  With the Complete topology, each
particle has $p$ neighbors.

In this paper we use these topologies as well as a parallel adaptation of the
Complete topology, called Random, that has been shown to approximate the
behavior of Complete with far less
communication~\citep{mcnabb-2009-large-particle-swarms}.  In the Random
topology, each particle randomly picks two other particles to share information
with at each iteration, along with itself.  Thus in both the Ring and the
Random topologies, all particles have three neighbors.

@@

Because the number of speculative evaluations depends on the number of
neighbors a particle has, the choice of topology is an important one.  The use
of the Complete topology, where every particle is a neighbor to every other
particle, would require $O(p^2)$ speculative evaluations per iteration.
Clearly it is much more desirable to have a sparse topology, where $O(np)$ is
much smaller than $O(p^2)$.  However, some functions are better optimized with
the Complete topology and the quick spread of information it entails than with
sparse topologies.  Accordingly, we use the Random topology described
in~\citep{mcnabb-2009-large-particle-swarms}, which has been shown to
effectively simulate the Complete topology.  In~\sec{results} we report the
results for performing speculative evaluation using both the Ring topology and
the Random topology on a number of common benchmark functions.

@@

In a centralized, or Master-Slave, parallel PSO algorithm, one machine, the
master, keeps track of all necessary information, and all other machines are
merely used to evaluate the objective function at various positions as directed
by the master~\citep{belal-2004-parallel-models-for-pso}.  To perform
speculative evaluation in such an architecture, the master generates the
positions to evaluate speculatively as in \eqref{eq:defpcasegen}.  After having
the slaves evaluate the objective function at all necessary positions, the
master then decides which position to accept for each particle, as in
\eqref{eq:val2updatelong}.  The outline of the procedure is given in
\alg{centralized}.

@@

In a distributed parallel PSO algorithm, individual processors not only perform
evaluations of particles, but also their movement.  The information for each
particle is not held by a central machine that directs the algorithm; instead,
each processor has the information for the particle or particles that it is in
charge of and must perform the steps of the algorithm for those
particles~\citep{mcnabb-2007-parallel-pso-using-mapreduce}.  Messages such as
values and positions for the neighborhood best are sent between processors.
There may still be some machine that collects information from all of the
particles and outputs the result of the algorithm, though that machine's
importance is much less than in centralized algorithms.

@@

While the idea of speculative evaluation in particle swarm optimization is
intriguing, the results in \sec{results} show that in many cases it requires
too many extra evaluations to produce competitive results.  On most functions,
the larger swarm size that can be used in na\"ive parallelizations leads to
better performance than is found by trying to do speculative evaluation to
exactly reproduce PSO.  However, if we keep the idea of speculative evaluation
while relaxing the requirement of exactly reproducing the behavior of the
original PSO algorithm, we see some impressive results.

@@

None of these methods fundamentally change the PSO algorithm.  They simply lead
to particles having different values for personal and neighborhood best
positions than would have occurred in standard PSO, because they receive
different information.  This kind of relaxation is fairly typical in the
parallelization of PSO~\citep{koh-2006-parallel-asynchronous-pso}.

@@

At this point it is also interesting to note a parallel between our methods and
parallel evolution
strategies~\citep{rudolph-1991-distributed-evolution-strategies}.  In evolution
strategies, a parent individual (representing a potential solution to some
objective function) produces a number of offspring by a mutation operator.  One
of the individuals is selected by a selection operator, and that individual
becomes the parent for the next
generation~\citep{beyer-2002-evolution-strategies}.  Our methods are similar,
where our mutation operator is simply the PSO motion equations and the
selection operator is either the indicator function introduced in \sec{sepso},
in the case of our original speculative algorithm, or the standard selector
operator based on fitness, in the case of this Pick Best technique.

@@

When not all of the branches are evaluated, there is some probability of not
evaluating the branch that was actually taken by the original particle.  There
are a few possibilities for recovery in this case.  One is to leave the
particle as it is, not accepting any of the speculative evaluations, because
none of them were correct.  This leads to particles being at different
iterations, as some particles guess correctly while others do not.  Thus we
lose exact compatibility with the original PSO, though this particular
relaxation is nothing new; asynchronous adaptations of PSO do the same
thing~\citep{koh-2006-parallel-asynchronous-pso}.  As an aside, it is equivalent
in this case to simply increment the iteration number of particles which fail
to correctly predict their branch.  This keeps the iteration number constant
across all particles, simplifying the work needed to be done in determining
neighbors when dynamic topologies are involved.  We call this technique Social
Promotion.

@@

We ran experiments to compare our speculative parallelization of PSO to the
na\"ive parallelization.  At each iteration of the algorithms, we use one
processor to perform one function evaluation for one particle, be it
speculative or non-speculative.  The speculative version actually performs
two iterations of PSO at each ``iteration,'' so we instead call each
``iteration'' a ``round of evaluations.''

Because our interest is in exploring the most efficient use of processors, we
must hold fixed the number of processors used at each round of evaluations in
all of our experiments, giving all approaches the same number of function
evaluations.  Thus while we label our graphs with ``Rounds of Evaluations,''
this is equivalent to the number of function evaluations used by each approach
(simply multiply by the number of processors to transform rounds of evaluations
to function evaluations).  We use ``Rounds of Evaluations'' to reflect
wall-clock time, as our primary interest is using extra processors in parallel
to improve performance on functions with long evaluation times.  A smaller
swarm running serial PSO would likely produce better results than we present in
the same number of function evaluations, but because the function evaluations
were not done in parallel it would take more wall-clock time.  When comparing
the performance of parallelized algorithms on long-running functions, we must
compare by wall-clock time with the same resources, not solely by the number of
function evaluations.

Using the same number of processors for each approach requires that our
speculative parallelizations have a smaller swarm size than the na\"ive
parallelization.  For each of the topologies we used, a particle has three
neighbors including itself.  As shown in Table~\ref{tab:evals}, this results in
$7$ speculative evaluations per particle.  With one evaluation needed for the
original, non-speculative particle, we have $8p$ evaluations for every two
iterations, where $p$ is the number of particles in the speculative swarm.  The
extra evaluations required in our speculative approach would instead be used to
evaluate particles in na\"ive parallelizations, so we compare swarms of size
$p$ in speculative evaluation with swarms of size $8p$ in na\"ive approaches.

Where the Complete topology would normally be preferred, we use a Random
topology in our speculative approach, as Complete leads to an explosion in the
number of speculative evaluations.  If speculative evaluation were not being
performed, it is possible that the Complete topology would be used.  However,
the Complete topology also requires a very large amount of interprocessor
communication in distributed PSO, so it is still quite possible that Random
would be used even with na\"ive parallelizations.  But, to be fair in our
comparisons, we compare to the na\"ive parallelization using both the Random
topology and the Complete topology (labeled PSO Random and PSO Complete in our
results).

We experimented with five common benchmark problems defined
in~\citep{herrera-2010-test-suite}: Rastrigin, Sphere, Schwefel 2.21, Griewank,
and Bohachevsky.

@@

The $c$ in the function definitions is a shifted center, in order to avoid
origin-seeking bias in the PSO algorithm and its
variants~\citep{monson-2005-origin-seeking-bias}.  We move the center of the
shifted functions (all except Bohachevsky) to be halfway between the center and
the boundary of the initialization region.  For example, if the initialization
region is $[-50,50]^D$, the center is $(25)^D$.  For all of these functions we
used 20 dimensions.

The parallelization techniques we compare are the na\"ive parallelization (here
labeled Na\"ive simply for ease of reference), our original speculative
approach (labeled SEPSO), and the four relaxations of SEPSO discussed in
\sec{relax}.  In presenting our results, we call the approach developed in
\sec{pickbest} Pick Best.  The methods described in \sec{pruning} through
\sec{wrong} are called Pick Best Pruned and Social Promotion Pruned, and the
method in \sec{manyiters} we call Many Iterations.

@@

Because pruning only requires two speculative evaluations per particle (along
with evaluating the original particle), we can use swarms of size
$\frac{1}{3}p$ to compare to a swarm of size $p$ with the na\"ive
parallelization, instead of $\frac{1}{8}p$ with other techniques.

@@

First we look at Rastrigin.  Rastrigin is a multi-modal function that is best
optimized with a large, Complete swarm.  It has been shown that with Rastrigin,
the more particles there are in the swarm, the lower function value it finds,
up to at least 4000 particles~\citep{mcnabb-2009-large-particle-swarms}.
Smaller swarms get caught in local optima.  Because our speculative algorithms
require significantly smaller swarm sizes, we would expect to not perform very
well on functions such as Rastrigin.  Our experiments show our intuition to be
correct.  In this experiment we used 240 processors, so SEPSO and Pick Best
each had 30 particles, and the Na\"ive algorithms had 240 particles.  As
expected, SEPSO and Pick Best converge quickly to worse local optima than
Na\"ive does.  \fig{rastrigin} shows the results graphically.

@@

Next we go to Sphere, the simplest of common benchmark functions.  The function
has no local optima (only a single global optimum) and is best optimized in
terms of function evaluations with a small swarm using a Complete topology.  We
expect our methods to be perfect for such functions, and our results show this
intuition again to be correct.  First we compare the basic SEPSO approach to
Na\"ive.  In \fig{basic-sphere} we can see that when both SEPSO and Na\"ive are
using a Random topology, SEPSO clearly outperforms Na\"ive.  However, the
comparison between SEPSO and Na\"ive Complete is not as encouraging---Na\"ive
Complete performs slightly better.

@@

When we introduce Pick Best, we see more impressive results.  As can be seen in
\fig{sphere-pickbest}, where SEPSO did not manage to perform as well as Na\"ive
Complete, Pick Best handily outperforms it with the same number of processors.

@@

Still more impressive with Sphere are the results of pruning and speculating
several iterations ahead.  Pick Best Pruned outperforms Pick Best, and Many
Iterations by far outperforms every other algorithm on this function.  Many
Iterations on average finds values that are more than 40 orders of magnitude
better than Na\"ive using the same number of processors.  This is shown in
\fig{sphere-manyiters}.  Social Promotion Pruned performed comparably with
SEPSO, and we do not clutter the graph with a line for it.

@@

Schwefel 2.21 is a function similar to Sphere, but Schwefel 2.21 is benefited
more by larger swarms than Sphere is.  Thus our speculative algorithms often
fail to perform as well as Na\"ive because of their smaller swarm size.  In
comparing SEPSO to Na\"ive in \fig{schwefel1}, we can see that SEPSO fails to
perform any better than Na\"ive with a Random topology, and it is by far
outperformed by Na\"ive with a Complete topology.

@@

As can be seen in \fig{schwefel2}, almost all of the speculative algorithms
perform worse than Na\"ive Complete, with Many Iterations performing comparably
to Na\"ive.  However, one of the speculative algorithms did perform better than
Na\"ive: Pick Best Pruned.  It seems that the larger swarm size available
because of pruning was the cause of the improved performance.  Note that Many
Iterations was initially performing comparably with Pick Best Pruned, but with
more iterations gradually degraded in performance to the level of Na\"ive
Complete.  It is likely that Many Iterations got caught in local optima, and
with more processors would have maintained better performance.

@@

Because the speculative algorithm got stuck with such a small swarm size, we
ran a set of experiments on Griewank with 800 processors, giving swarms of size
100 and 800 instead of 30 and 240.  \fig{basic-griewank2} shows the results.
\fig{basic-griewank2} is very similar to \fig{basic-griewank1}, except that
speculative evaluation finds the global optimum essentially 100\% of the time.
Thus with a few more processors, our algorithm finds the optimum on average
twice as fast as the Na\"ive approach without a significant loss of accuracy.

@@

In order to show the premature convergence properties of the various approaches
we have presented, we will show results first with swarms using 240 processors,
where these are more visible.  Then we will present an overall result for 800
processors, where using our techniques is generally more successful.
